Efficient inference using the approximate Riemannian geometry of stochastic population models

### 2013-12-16 paper/paper.tex
Merging introduction to paper

\author{Ben Calderhead\affil{1}{University College London, London, United Kingdom}, Colin S. Gillespie\affil{2}{Newcastle University, Newcastle upon Tyne, United Kingom}}
% \email{b.calderhead@ucl.ac.uk}
% \email{colin.gillespie@newcastle.ac.uk}
% \\
% 	   Centre for Mathematics and Physics in the Life Sciences and Experimental Biology,\\University College London, UK
% 	   \and 
% 	   School of Mathematics \& Statistics, Newcastle University,
% 	   NE1 7RU, UK
\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

@@

\begin{abstract} -- enter abstract text here -- \end{abstract}

@@

@@

\begin{itemize}
\item Inference for MJP is difficult: non-linear likelihood, computational
  expensive, proposals expensive, analytically intractable likelihood
\item Data: Partially observed discrete, noisy time-course.
\item Existing work for likelihood evaluations: 
\begin{itemize}
\item Likelihood calculations
\begin{itemize}
\item Exact: Boys et al, slow, not noise
\item Approx SDE - Andy and Darren (with Noise)
\item Approx MC - Andy and Colin (no noise)
\item Approx LNA - Mark (no noise), Kolmo' (no noise, but check)
\end{itemize}
\end{itemize}
\item Novel stuff
\end{itemize}
Different from Mark's LNA paper:
\begin{itemize}
\item Analysis of impact of the approximate proposal scheme
\item Using Noise
\item Using delayed acceptance scheme so exact inference
\item Time course data
\item Using real data
\end{itemize}

Importance of stochastic population models for further scientific understanding of natural systems vs deterministic models.  For example in ecology, where we often have to deal with discrete noisy data.  Importance of parametric sensitivity has already been shown for determining vital properties of biological systems such as robustness and also in experimental design where the sensitivity gives information about the identifiability of individual rate constants.

@@

  \begin{itemize}
  \item Calculate exact proposal for comparison (\textbf{17/8: CSG})
  \item Construct LNA model for ID (\textbf{14/9: CSG/BC})
  \item Graphs:
    \begin{itemize}
    \item Data: \textbf{17/8: CSG}
    \item Geometry proposal: 3 proposals for the 3 derviates: k1\^2, k1k2, k2\^2
      \textbf{24/8: }
    \item Likelihood: Exact likelihood vs LNA vs MC \textbf{24/8: }
    \item Delay acceptance(?)
    \item Sampling efficiences M-H vs RM (appendix)
    \end{itemize}
  \end{itemize}

@@

    \begin{itemize}
    \item Construct moment equations (\textbf{17/8: CSG}) Lognormal?
    \item Construct proposal
    \item Delay acceptance
    \item Run experiments
    \item Graphs:
      \begin{itemize}
      \item Data
      \item Sampling efficiences M-H vs RM (appendix)
      \item Posterior compare to Stumpf
      \end{itemize}
    \end{itemize}

@@

36 parameters - 18 birth, 18 death. 3 water treatments, 3 nitrogen treatments
and 3 blocks, with interactions. 5*27 (I think) latent data sets. Exploit Sub-mannifolds for choosing parameter sets
Poisson error structure.

    \begin{itemize}
    \item Construct moment equations (appendix) \textbf{Done}
    \item Construct proposal (appendix)
    \item Delay acceptance
    \item Run experiments
    \item Graphs:
      \begin{itemize}
      \item Data \textbf{Done}
      \item Sampling efficiences M-H vs RM; ESS
      \item Posteriors
      \end{itemize}
    \end{itemize}

@@

\begin{itemize}
\item Use of geometry vital in tackling real problems
\item Could also use other approximation schemes such as the LNA??
\end{itemize}

%% == end of paper:

@@

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[]{../graphics/figure1}}
\end{center}
\caption{Stochastic simulation of the immigration-death process, with parameters
$k_1=k_2=1$ and $n(0)=0$. The black line shows the full stochastic realisation, 
the red dots indiciate the observed data.}
\end{figure}
\begin{figure*}[ht]
\begin{center}
\centerline{\includegraphics[width=.7\textwidth]{../graphics/figure2}}
\caption{Comparison of the exact and the moment closure expected fisher information matrix. \textit{Note: I'm not happy with the layout, still working on that}.}\label{afoto2}
\end{center}
\end{figure*}
% 
% \begin{figure*}
% \centering
% \includegraphics[]{../graphics/figure1}
% \caption{blah}
% \end{figure*}

>>>>

\author{Ben Calderhead\affil{1}{University College London, London, United
    Kingdom}, Colin S. Gillespie\affil{2}{Newcastle University, Newcastle upon
    Tyne, United Kingdom}}
% \email{b.calderhead@ucl.ac.uk}
% \email{colin.gillespie@newcastle.ac.uk}
% \\
% 	   Centre for Mathematics and Physics in the Life Sciences and Experimental Biology,\\University College London, UK
% 	   \and 
% 	   School of Mathematics \& Statistics, Newcastle University,
% 	   NE1 7RU, UK
\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

@@

\begin{abstract}
We introduce an efficient and novel approach for Bayesian statistical
inference of stochastic population models with complex dynamics,
unobserved species and large numbers of parameters. Despite their
increasing ubiquity, Bayesian analysis of such stochastic models still
presents major challenges, as the exact likelihood is often computa-
tionally intractable for realistically sized models and full exploration
of the parameter space may be inhibited by very strong correla-
tion structure. Our methodology exploits the natural underlying ap-
proximate Riemannian geometry of the model parameters induced
by estimating the likelihood using moment closure techniques. We
demonstrate how this geometric approach allows us to perform in-
ference in a statistically and computationally efficient manner, and
in doing so we highlight the link between the Riemannian structure
of the parameters and sensitivity analysis of the dynamical system.
We illustrate the performance of our approach using synthetic ex-
amples, as well as a realistically-sized partially observed ecological
model with 50 unknown parameters, with data describing the evolu-
tion of aphid populations. For the examples in this paper, efficient
computer code written in Matlab and C is available from the authors
upon request.
\end{abstract}

@@

The use of stochastic models is rapidly becoming ubiquitous in many areas of the
natural sciences, where it is necessary to model the intrinsic stochasticity of
a dynamical system in or- der to fully characterise its main modes of behaviour
[REFs]. Markov Jump Process models - what are they and why are they useful?

Mathematical approaches are essential for making sense of these complex models;
well developed for deterministic mod- els, such as ordinary differential
equations, but still underde- veloped for stochastic models.

Why? Likelihood is often computationally intractable for
all but the simplest of Markov Jump Process models.
Sensitivity analysis plays an important role in elucidating
the behaviour of mathematical models with complex dynam
ics - how do variations in parameters accpet the output? Only
fairly recently have methods been published for estimating the
local sensitivity information for a stochastic chemical kinetic
model via use of the linear noise approximation [REFs].
local versus global sensitivity analysis

Sensitivity information at the maximum likelihood is
based on a linearisation around this point and is thus a lo-
cal measure. Given the strong nonlinearities present in many
current models, this provides limited information about the
dynamics of the system.

FIM is often used to describe the local sensitivity at the
maximum likelihood however this makes strong assumption
of normality around that set of parameters, which may be
inappropriate for strongly nonlinear systems.

Bayesian methodology in particular offers a consistent,
probabilistic framework for reasoning under uncertainty,
which may occur both in the data and the model structure
itself.

Posterior distribution as a global measure of sensitivity of
the system. However drawing samples from this is generally
very challenging.

FIM provides a link between local and global sensitivity
analysis.

FIM is usually used to measure the reliability of inferences
around the maximum likelihood and is often understood ge-
ometrically as the curvature of the log-likelihood around this
point. However, the expected Fisher Information in fact gives
us much more geometric information than this; it defines a
metric tensor at any point of parameter space and thus pro-
vides a natural representation of the parameter space as a
Riemannian manifold, where distances between two sets of
parameters are defined in terms of the relative changes they
induce on the likelihood.

We use a moment closure technique to obtain local ap-
proximations to the likelihood at the current point - such an
approach extends the LNR as we may now choose the pdf
with which to approximate the local likelihood. We demon-
strate how the expected Fisher Information may be obtained
from this moment closure approximation and that this forms
a Riemannian manifold that approximates the local sensitiv-
ity of the model at each possible set of parameters. Further,
we demonstrate the utility of this using differential geometric
MCMC algorithms that allow us to sample from the poste-
rior distribution of the parameters of stochastic population
models, by utilising the local sensitivity information at each
point. Essentially a locally adaptive algorithm that exploits
the local sensitivity at each point, allowing us to move from
local to global sensitivity information in a computationally
and statistically efficient manner.

@@

\begin{itemize}
\item Inference for MJP is difficult: non-linear likelihood, computational
  expensive, proposals expensive, analytically intractable likelihood
\item Data: Partially observed discrete, noisy time-course.
\item Existing work for likelihood evaluations: 
\begin{itemize}
\item Likelihood calculations
\begin{itemize}
\item Exact: Boys et al, slow, not noise
\item Approx SDE - Andy and Darren (with Noise)
\item Approx MC - Andy and Colin (no noise)
\item Approx LNA - Mark (no noise), Kolmo' (no noise, but check)
\end{itemize}
\end{itemize}
\item Novel stuff
\end{itemize}
Different from Mark's LNA paper:
\begin{itemize}
\item Analysis of impact of the approximate proposal scheme
\item Using Noise
\item Using delayed acceptance scheme so exact inference
\item Time course data
\item Using real data
\end{itemize}
Why MC is better than LNA
\begin{itemize}
\item Higher-order approximation - REF: J CHEM Phy
\item Use other closure distributions: Lognormal - beta-bionomial
\item REF other recent MC PNAS papers - Kazinesis and Zurich paper
\end{itemize}

Importance of stochastic population models for further scientific understanding of natural systems vs deterministic models.  For example in ecology, where we often have to deal with discrete noisy data.  Importance of parametric sensitivity has already been shown for determining vital properties of biological systems such as robustness and also in experimental design where the sensitivity gives information about the identifiably of individual rate constants.

@@

  \begin{itemize}
  \item Calculate exact proposal for comparison (\textbf{17/8: CSG})
  \item Graphs:
    \begin{itemize}
    \item Data:
    \item Geometry proposal: 2 proposals for the 3 derivatives: k1\^2, k1k2, k2\^2
    \item Likelihood: Exact likelihood vs MC
    \item Sampling efficiencies M-H vs RM (appendix)
    \end{itemize}
  \end{itemize}

@@

    \begin{itemize}
    \item Construct moment equations:  Lognormal or Beta Binomial
    \item Construct proposal
    \item Run experiments
    \item Graphs:
      \begin{itemize}
      \item Data
      \item Sampling efficiencies M-H vs RM (appendix)
      \end{itemize}
    \end{itemize}

@@

36 parameters - 18 birth, 18 death. 3 water treatments, 3 nitrogen treatments
and 3 blocks, with interactions. 5*27 (I think) latent data sets. Exploit Sub-manifolds for choosing parameter sets
Poisson error structure.

    \begin{itemize}
    \item Construct moment equations (appendix) \textbf{Done}
    \item Construct proposal (appendix)
    \item Run experiments
    \item Graphs:
      \begin{itemize}
      \item Data \textbf{Done}
      \item Sampling efficiencies M-H vs RM; ESS
      \item Posteriors
      \end{itemize}
    \end{itemize}

@@

\begin{itemize}
\item Use of geometry vital in tackling real problems
\end{itemize}
%% == end of paper:

@@

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[]{../graphics/figure1}}
\end{center}
\caption{Stochastic simulation of the immigration-death process, with parameters
$k_1=k_2=1$ and $n(0)=0$. The black line shows the full stochastic realisation, 
the red dots indicate the observed data.}\label{F1}
\end{figure}
\begin{figure*}[ht]
\begin{center}
\centerline{\includegraphics[width=.7\textwidth]{../graphics/figure2}}
\caption{Comparison of the exact and the moment closure expected fisher information matrix. }
\label{F2}
\end{center}
\end{figure*}
% 
% \begin{figure*}
% \centering
% \includegraphics[]{../graphics/figure1}
% \caption{blah}
% \end{figure*}
